---
title: "Getting Started with historicalBorrow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with historicalBorrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set TRUE when CmdStan is available in the build environment
)
```

## Overview

The **historicalBorrow** package provides a simple, high-level interface for
incorporating historical clinical trial data into Bayesian analyses. Under the
hood, Stan models are generated, compiled, and run automatically — no Stan
coding is required.

The four-step workflow is:

| Step | Function | Purpose |
|---|---|---|
| 1 | `prepare_data()` | Validate and structure data |
| 2 | `fit_map()` | Estimate MAP prior from historical studies |
| 3 | `robustify_map()` | Mix MAP with vague prior (RMAP) |
| 4 | `fit_borrowing_model()` | Analyse current trial with borrowing |

## Installation

```{r install, eval=FALSE}
# Install the Stan backend first
install.packages("cmdstanr",
  repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()

# Install historicalBorrow
# devtools::install_github("Teniola17/historicalBorrow")
```

```{r load}
library(historicalBorrow)
```

---

## Binary Outcome Example

### 1 — Prepare data

`example_trial_data` ships with the package and contains simulated binary,
continuous, and count trial data (6 historical studies + 1 current trial each).

```{r binary-prep}
pd <- prepare_data(
  hist_data    = example_trial_data$binary$hist,
  current_data = example_trial_data$binary$current,
  outcome      = "binary"
)
print(pd)
```

### 2 — Fit MAP prior

The MAP prior is estimated from the historical studies using a Bayesian
hierarchical binomial-logit model. The posterior predictive distribution for a
new exchangeable study (`mu_pred`) is approximated as a K-component mixture of
normals (K chosen by BIC).

```{r binary-map}
set.seed(2024)
mp <- fit_map(
  data          = pd,
  chains        = 4L,
  iter_warmup   = 1000L,
  iter_sampling = 2000L
)
print(mp)
plot(mp)   # histogram of mu_pred draws + mixture PDF overlay
```

The `ess_prior` field reports the approximate equivalent historical sample size:
how many current-trial patients the MAP prior is worth.

### 3 — Robustify

Add a vague component (weight 20%) to protect against prior-data conflict.

```{r binary-rmap}
rmp <- robustify_map(mp, weight = 0.8)
print(rmp)
plot(rmp)  # MAP vs vague vs total RMAP density
```

### 4 — Fit borrowing model

```{r binary-fit}
bf <- fit_borrowing_model(
  current_data = pd,
  prior        = rmp,
  model        = "rmap"
)
print(bf)
summary(bf)
```

### Diagnostics

```{r binary-diag}
diag <- diagnostics(bf)
print(diag)
diag$trace_plots   # ggplot trace plots for main parameters
```

Warnings are surfaced automatically when R-hat > 1.05, divergences are
detected, or prior-data conflict is flagged.

### Visualisations

```{r binary-plots}
# Prior and posterior density overlay
plot_prior_posterior(bf)

# Credible interval comparison (with vs. without borrowing)
plot_borrowing_effect(bf)

# Shrinkage of posterior towards the MAP prior
plot_shrinkage(bf)

# Via the unified plot() dispatch
plot(bf, type = "prior_posterior")
plot(bf, type = "effect")
plot(bf, type = "shrinkage")
plot(bf, type = "trace")
```

### Go/No-Go decision

```{r binary-decision}
dc <- decision_criteria(
  fit                   = bf,
  theta_threshold       = 0.25,   # minimum clinically meaningful response rate
  direction             = "greater",
  probability_threshold = 0.80
)
print(dc)
```

---

## Continuous Outcome Example

```{r continuous}
pd_c  <- prepare_data(
  example_trial_data$continuous$hist,
  example_trial_data$continuous$current,
  outcome = "continuous"
)
mp_c  <- fit_map(pd_c, seed = 1)
rmp_c <- robustify_map(mp_c, weight = 0.8)
bf_c  <- fit_borrowing_model(pd_c, prior = rmp_c, model = "rmap")
summary(bf_c)
plot_prior_posterior(bf_c)
```

---

## Count Outcome Example

```{r count}
pd_n  <- prepare_data(
  example_trial_data$count$hist,
  example_trial_data$count$current,
  outcome = "count"
)
mp_n  <- fit_map(pd_n, seed = 1)
bf_n  <- fit_borrowing_model(pd_n, prior = mp_n, model = "map")
summary(bf_n)
diagnostics(bf_n)
```

---

## Advanced: Power Prior

The power prior scales the historical log-likelihood by a factor `a0 ∈ [0, 1]`.
Setting `a0_fixed = -1` estimates `a0` from the data (modified power prior).

```{r power}
bf_power <- fit_borrowing_model(
  current_data = pd,
  prior        = NULL,
  model        = "power",
  hyperpriors  = list(
    a0_fixed = -1,     # estimate a0
    a0_alpha = 1,      # Beta(1,1) hyperprior on a0
    a0_beta  = 1
  )
)
summary(bf_power)
# The a0 posterior tells you how much the data supports borrowing:
# a0 near 1 => current and historical data are consistent
# a0 near 0 => conflict detected, borrowing suppressed
```

---

## Advanced: Commensurate Prior

The commensurate prior links the current and historical treatment parameters
via a precision parameter `tau_comm`. A heavy-tailed Cauchy(0, 1) hyperprior
on `tau_comm` allows the data to choose the borrowing strength automatically.

```{r commensurate}
bf_comm <- fit_borrowing_model(
  current_data = pd,
  prior        = NULL,
  model        = "commensurate"
)
summary(bf_comm)
# Small tau_comm posterior => strong borrowing (data are commensurate)
# Large tau_comm posterior => weak borrowing (prior-data conflict)
```

---

## Object Classes

| Class | Created by | Key slots |
|---|---|---|
| `bb_data` | `prepare_data()` | `outcome`, `hist_data`, `stan_data_map`, `stan_data_current` |
| `map_prior` | `fit_map()` | `mixture`, `mu_pred_draws`, `ess_prior`, `draws` |
| `rmap_prior` | `robustify_map()` | inherits `map_prior` + `mixture_rmap`, `ess_prior_rmap` |
| `borrowing_fit` | `fit_borrowing_model()` | `draws`, `prior`, `model`, `stan_code` |
| `hb_diagnostics` | `diagnostics()` | `rhat`, `divergences`, `prior_data_conflict`, `trace_plots` |
| `decision_result` | `decision_criteria()` | `go`, `probability`, `posterior_ci` |

---

## Session Info

```{r session}
sessionInfo()
```
