---
title: "Time-Dependent Priors (TDP)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Dependent Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

The **Time-Dependent Prior** (TDP) extends historical borrowing to account for
temporal trends and drift in treatment effects. It uses an Ornstein-Uhlenbeck
continuous-time process to model how the treatment parameter evolves over time,
with automatic down-weighting of older studies via a correlation decay parameter.

### Key Differences from MAP

| Feature | MAP | TDP |
|---|---|---|
| Historical studies | Treated as exchangeable | Ordered by time |
| Borrowing weight | Uniform across studies | Decays with time gap |
| Control of borrowing | Fixed prior weight | Automatic via OU process + `rho` |
| When to use | Studies from similar era | Studies across decades/years |

---

## Quick Start

### Step 1: Prepare data with times

```{r tdp-prep}
# Add a 'time' column to historical data (study year, trial date, etc.)
hist_with_times <- example_trial_data$binary$hist
hist_with_times$time <- c(2010, 2012, 2014, 2015, 2016, 2017)

pd <- prepare_data(
  hist_data    = hist_with_times,
  current_data = example_trial_data$binary$current,
  outcome      = "binary"
)
```

Alternatively, pass times separately:

```{r tdp-prep-alt}
pd <- prepare_data(
  hist_data    = example_trial_data$binary$hist,
  current_data = example_trial_data$binary$current,
  outcome      = "binary"
)
# Times will be passed to fit_tdp() below
```

### Step 2: Fit the time-dependent prior

```{r tdp-fit}
set.seed(2024)
tdp <- fit_tdp(
  data         = pd,
  hist_times   = c(2010, 2012, 2014, 2015, 2016, 2017),
  current_time = 2024,
  chains       = 4L
)
print(tdp)
plot(tdp)
```

The posterior of `rho` controls the decay:
- **`rho` near 1**: Slow decay, older studies retain influence
- **`rho` near 0**: Fast decay, only recent studies matter
- **Automatic**: Data-driven via Beta(2,2) hyperprior on `rho`

### Step 3: Robustify (optional)

```{r tdp-robust}
rtdp <- robustify_tdp(tdp, weight = 0.8)
print(rtdp)
plot(rtdp)
```

### Step 4: Use in borrowing model

```{r tdp-borrowing}
bf_tdp <- fit_borrowing_model(
  current_data = pd,
  prior        = rtdp,
  model        = "rtdp"
)
summary(bf_tdp)
plot_prior_posterior(bf_tdp)
```

---

## Understanding the Time-Dependent Prior Model

The TDP uses an **Ornstein-Uhlenbeck process** parameterised as:

$$\theta_h | \theta_{h-1} \sim N(\mu + \rho^{\Delta t_h}(\theta_{h-1} - \mu), \sigma^2 \frac{1-\rho^{2\Delta t_h}}{1-\rho^2})$$

where:
- **$\theta_h$** = treatment effect (logit scale for binary, natural scale for continuous, log-rate for count)
- **$\mu$** = long-term mean effect
- **$\rho$** = correlation decay per unit time (Beta(2,2) prior)
- **$\sigma$** = temporal noise SD
- **$\Delta t_h$** = time gap between studies $h-1$ and $h$

### Parameters Estimated

| Parameter | Role | Prior |
|---|---|---|
| `mu` | Long-term mean | N(0, 2.5) [binary/count]; N(0, 100) [continuous] |
| `mu1` | Initial effect (first study) | N(mu, 2.5) [binary]; N(mu, 50) [continuous] |
| `sigma1` | Initial SD | N(0, 1) [binary]; N(0, 10) [continuous] |
| `tau` | Temporal noise | N(0, 0.5) [binary]; N(0, 10) [continuous] |
| `rho` | Decay per unit time | Beta(2, 2) [symmetric, moderate] |
| `theta[h]` | Each study's effect | From OU evolution |

### Prediction (Generated Quantity)

At current time $t^*$:

$$\theta^* \sim N(\mu + \rho^{\Delta t^*}(\theta_H - \mu), \sigma^2 \frac{1-\rho^{2\Delta t^*}}{1-\rho^2})$$

where $\Delta t^* = t^* - t_H$ (time since the latest historical study).

---

## Example: Drift in Efficacy

Suppose treatment efficacy has improved over time (due to better patient selection,
dosing refinement, etc.). The TDP automatically down-weights the old, less
effective historical studies:

```{r tdp-drift}
# Historical studies span 2005 to 2020 with improving efficacy
# Current trial in 2024 — wants more recent evidence

# Using TDP with rho ~ Beta(2,2):
#   - Studies from 2020: full influence
#   - Studies from 2015: ~70% influence (5-year decay)
#   - Studies from 2005: ~10% influence (19-year decay)

# Using MAP (ignoring time):
#   - All studies treated equally, regardless of age
#   - May underestimate current efficacy if trend is positive

tdp_modern <- fit_tdp(pd, hist_times = c(2005, 2010, 2013, 2015, 2018, 2020),
                      current_time = 2024)
print(tdp_modern)
# The rho posterior will be pulled toward higher values (slower decay)
# if the trend is detected in the data
```

---

## Advantages & Limitations

### Advantages

✓ **Automatic down-weighting** of older data based on time and correlation decay
✓ **Data-driven borrowing** — the extent of borrowing is learned from the data
✓ **Realistic for long-term trials** — accounts for drift in treatment effects
✓ **Interpretable** — `rho` posterior tells you about temporal trends

### Limitations

✗ **Requires time information** — all historical studies must have a date/time
✗ **Assumes continuous process** — may not suit discrete "eras" of treatment
✗ **Computational cost** — slightly more parameters than MAP (one `rho` per model)
✗ **Sensitive to time spacing** — uneven gaps between studies can affect OU evolution

---

## Robustifying the TDP

Just as MAP can be robustified, so can TDP:

```{r tdp-mix}
# Pure TDP: 100% weight on temporal evolution
bf_tdp_pure <- fit_borrowing_model(pd, prior = tdp, model = "tdp")

# Robust TDP: 80% TDP + 20% vague prior
# Protects against prior-data conflict (e.g., if the current trial
# has a dramatically different effect than the trend predicts)
rtdp <- robustify_tdp(tdp, weight = 0.8)
bf_tdp_robust <- fit_borrowing_model(pd, prior = rtdp, model = "rtdp")

# Compare
summary(bf_tdp_pure)
summary(bf_tdp_robust)
```

The vague component has the same tuning as for MAP:

```{r tdp-vague}
rtdp_wide <- robustify_tdp(
  tdp,
  weight      = 0.7,  # Only 70% TDP
  vague_prior = list(mean = 0, sd = 20)  # Very wide vague prior
)
```

---

## Session Info

```{r}
sessionInfo()
```
